Synthesising Complex Termination Arguments

### 2014-06-12 papers/termination/paper.tex
messy stuff

@@

\section{Related Work}
Primarily divided into structural (syntactic) vs. semantic.  For example, terminationg of term rewriting systems
is largely structural, whereas DWF analysis (cf. Terminator) is more semantic -- it just deals
with transition relations, but uses some amount of syntax in that it identifies loops \& looping
paths.

\section{Motivational Examples}
A non-linear program with a linear ranking function (+ program with non-linear operations):
\begin{lstlisting}
while (x > 0)
 x = (x - 1) & x;
\end{lstlisting}

A linear program with a non-linear ranking function:
\begin{lstlisting}
while (x != 0)
 x = -x / 2;
\end{lstlisting}

@@

>>>>

%completeness without disj well-foundedness
%. only non-strict inequalities can be transformed using Farkas’ Lemma

@@

\section{Related Work}
\subsection{Termination analysis}
The existing work on termination is primarily divided into structural (syntactic) vs. semantic approaches.  
For example, terminationg of term rewriting systems is largely structural, whereas DWF analysis (cf. Terminator) is more semantic. While it deals
with transition relations, it also uses some amount of syntax in that it identifies loops \& looping paths.

A program is lasso-shaped if it is composed by a stem followed by a single loop without branching, i.e. there is only one path. 
Consequently, the termination argument can be very simple. There are numerous techniques specialised in
proving termination of lasso-shaped programs efficiently \cite{DBLP:conf/popl/Ben-AmramG13,DBLP:conf/cav/BradleyMS05,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/P04}.

Cook et al. are closer to a semantic approach by viewing the termination proving task as a binary reachability analysis \cite{DBLP:conf/pldi/CookPR06}.
%Terminator incrementally constructs the union of ranking relations.
%Given that the this union is not well-founded (it is only disjunctively well-founded), 
%Terminator must prove that the binary reachability relation of the program (the transitive closure of the transition relation restricted to reachable states) 
%is contained in the union of ranking relations.
%Thus, this approach requires an evolved binary reachability, and, implicitly, the assistance of a safety prover.
In particular, their technique incrementally constructs the union of ranking relations.
Given that the this union is not well-founded (it is only disjunctively well-founded),
they must show that, for every sequence, one of the ranking functions decreases
between the pre- and post-state. %In other words: it 
%must find all pairs of states reachable from the program's initial state 
%such that one is reachable from the other and it must show 
%that the value of one of the ranking functions decreases from s1 to s2. 
Consequently, most of the analysis time is spent in reachability analysis that is to check if the current
transition invariant reached a fixed point.

As we want to avoid the requirements of a binary reachability analysis, our termination argument is not a set
of ranking functions, but a single ranking function, which we prove is enough for proving the termination of any termination program. 
With a single ranking function we only need to show that the rank decreases from the
pre- to post-state after executing each single transition step.
The checking process is therefore easy. For the construction of the ranking function, we 
formulate the problem as a second order satisfaction formula and we solve it via program synthesis.

Lee et al. make use of transition predicate abstraction, algorithmic learning, and decision procedures to compute transition
invariants as proofs for the termination of linear programs \cite{DBLP:conf/cav/LeeWY12}.

There is very limited work on generating ranking functions for relations over machine integers.
Cook et al. present a method based on a reduction to Presburger
arithmetic, and a template-matching approach for predefined classes of
ranking functions based on reduction to SAT- and QBF-solving \cite{DBLP:conf/tacas/CookKRW10}.

Leike and Heizmann present a new method for the constraint-based synthesis
of termination arguments for linear loop programs based on
linear ranking templates \cite{DBLP:conf/tacas/LeikeH14}.

Learning: \cite{DBLP:journals/corr/HeizmannHP14}

Polynomial programs: \cite{DBLP:conf/vmcai/BradleyMS05}

%-- Usually the termination argument for the program LOOPS
%on Figure 1 is based on a lexicographic combination
%of well-founded orderings.

%-- the union T of ranking relations, however, is in general not well-founded
%(it is only disjunctively well-founded in the terminology of [23]). This is
%why it would not be sufficient to show the inclusion R \subseteq  T, and why we must in stead prove R_I^+ \subseteq T.

\subsection{Synthesis}

%% ---
%% constructs and checks the termination argument

%% T ERMINATOR’s most distinguishing aspect, with respect to pre-
%% vious methods and tools for proving program termination, is how it
%% shifts the balance between the two tasks of constructing and respec-
%% tively checking the termination argument. The classical method is
%% to construct an expression defining the rank of a state and then to
%% check that its value decreases in every transition from a reachable
%% state to a next one. The construction of the ranking function is the
%% hard part and forms a task that needs to be applied to the whole
%% program. The checking part is relatively easy. In our method, the
%% task of constructing ranking functions is the relatively easy part;
%% they are constructed on demand based on the examination of only
%% a few selected paths through the program.

%% Previously, it was not known whether binary reachability analysis could be made
%% practical. The challenge raised by our approach was to show that
%% this is indeed the case.

%% ----------
%% Efficient synthesis of ranking functions for machine-level bit-vectors, how-
%% ever, has remained an open problem.
%% ---------

\section{Motivational Examples} 
We next discus a few examples that illustrate the difficulties in termination checking for low-level code. In particular, they denote situations that arise 
frequently in low-level code and are not handled by existing techniques.

%We now present a few examples that justify

\subsection{A non-linear program with a linear ranking function.}
The first example we consider is a program with non-linear operations: 
\begin{lstlisting}
while (x > 0)
 x = (x - 1) & x;
\end{lstlisting}

In order to find a ranking function for this example, it is necessary to take into account
the semantics of the bit-wise AND operator, which is not easily done when working with mathematical integers \cite{}.
Our technique finds that a possible ranking function is the linear function
$R(x) = x$, whose value decreases with
every iteration, but it can not decrease indefinitely as it is bounded from below.

\subsection{A linear program with a non-linear ranking function.}
\begin{lstlisting}
while (x != 0)
 x = -x / 2;
\end{lstlisting}

\subsection{Differences in the termination behaviour for reals and floats.}
\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}

\begin{lstlisting}
while (x > 0.0) x *= 0.5;
\end{lstlisting}

\subsection{A linear program with lexicografic non-linear function.}
\begin{lstlisting}
while ( q >= 0 ) {
  q = q - y;
  y = y + 1;
}
\end{lstlisting}

Every execution of the above program can be partitioned into two phases: in the first phase $y$ increases
until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
%Depending on the initial values of y and q, either phase might be skipped
%altogether.
\cite{DBLP:conf/tacas/LeikeH14} proposes using a linear ranking template in order to obtain a ranking function that proceeds
through a fixed number of phases in the program execution. Each phase is
ranked by an affine-linear function, and ends when this function becomes non-positive.

Lexicographic non-linear ranking functions consist of lexicographically ordered components
of non-linear functions. As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
loop transition leads to a decrease with respect to the lexicographic
ordering for this tuple. Therefore no function may increase unless a function of
a lower index decreases. Additionally, at every step, there must be at least one
function that decreases.

\subsection{Lexicographic ranking function with strict inequalities.}
Approaches based on Farkas’ Lemma can only handle non-strict inequalities \cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/vmcai/P04}.
[2,5,6,7,14,20,24,25

@@

Our method for ranking function synthesis can be stated as follows:
discuss what spec is used (non-lexicographic vs lexicographic) + the completeness claims.
Any termination guarantees?
